If the Null Fits, You Must Omit: Ubiquitous False Detections of Group Polarization

### 2025-08-14 GroupPolarizationStatmod/paper/ms.tex
added date
fixed error in algorithm logic

To find examples of plausibly spurious group polarization we guess at what
latent mean, combined with a latent pre- and post-deliberation 
latent standard deviations, could generate it. We calculate the observed group
polarization for these latent parameters by transforming the latent pre- and
post-deliberation distributions from continuous normal distributions to ordinal
"observations" by integrating over each ordinal bin, then calculating the observed
mean for each experiment *phase*, i.e., both the pre- and post-deliberation phases. 
If the null model can generate group polarization in this case, then the detection
is labelled *plausibly false*, i.e., *plausibly spurious*, and we call the 
generating latent parameters the *plausible null parameters*.

To evaluate whether the plausible null parameters we identify could realistically
fool a significance test, we implemented a generative simulation of each plausibly
false detection of group polarization and calculated the *family-wise error rate
(FWER)* and the *false discovery rate (FDR)*. We use the published number of
participants $N$ for each of the 54 conditions that yielded plausibly false
detections of group polarization. The FWER is also known as the Type I
error, or the false positive rate: it is the probability that a detection of an
effect occurs given that in reality there was a null effect.  The FDR is the
probability that a published positive detection of an effect is a false positive,
i.e., a detection of a null effect. High FWER and FDR for an experimental condition
indicates that its design is weak, since we identify only one of perhaps infinitely
many latent parameter combinations that can generate spurious group polarization.

We use this setup to demonstrate how *adversarial* style testing of experimental
designs with counterfactual null models can be used to estimate important
metascience parameters. In this study we use our FWER calculations to estimate
the minimum Cohen's $d$ that should be used to determine whether an effect is
``significant'' or not that will limit the FWER to 5\%. In the Discussion we
sketch how our model can then be extended to rigorously calculate statistical power,
i.e., the probability that a significant effect is detected given that there 
is indeed a significant effect.

@@

    \State $\Delta \gets \meanobs(\mu, \sigma) - \meanobsemp$ 
      \Comment{Initialize simulation error.}
    \While {$\Delta^2 > \epsilon$ and $i < i_\mathrm{max}$} \Comment{Search until desired
    accuracy or max iterations reached.}
      \State $\sigma' \sim \mathcal{N}(\sigma, \delta)$ \Comment{Draw new variance to test.}
      \State $\Delta' \gets \bar{o}(\mu, \sigma') - \bar{\mathbf{o}}$ 
        \Comment{Calculate error using new variance.}
      \If {$\Delta' - \Delta < 0$}
        \State $\sigma \gets \sigma'$
          \Comment{Update $\sigma$ if the error is less than before.}
      \EndIf
      \State $\Delta \gets \bar{o}(\mu, \sigma) - \bar{\mathbf{o}}$
        \Comment{Update simulation error if $\sigma$ is updated.}
    \EndWhile \\

>>>>

(METHODS OVERVIEW)

@@

    \State $\Delta \gets \meanobs(\mu, \sigma) - \meanobsemp$ 
      \Comment{Initialize simulation error.}
    \While {$\Delta^2 > \epsilon$ and $i < i_\mathrm{max}$} \Comment{Search until desired
    accuracy or max iterations reached.}
      \State $\sigma' \sim \mathcal{N}(\sigma, \delta)$ \Comment{Draw new variance to test.}
      \State $\Delta' \gets \bar{o}(\mu, \sigma') - \bar{\mathbf{o}}$ 
        \Comment{Calculate error using new variance.}
      \If {$\Delta' - \Delta < 0$}
        \State $\sigma \gets \sigma'$
          \Comment{Update $\sigma$ if the error is less than before.}
        \State $\Delta \gets \bar{o}(\mu, \sigma) - \bar{\mathbf{o}}$
          \Comment{Update simulation error if $\sigma$ is updated.}
      \EndIf
    \EndWhile \\
